---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# HMMERutils

<!-- badges: start -->
[![GitHub issues](https://img.shields.io/github/issues/currocam/HMMERutils)](https://github.com/currocam/HMMERutils/issues)
[![GitHub pulls](https://img.shields.io/github/issues-pr/currocam/HMMERutils)](https://github.com/currocam/HMMERutils/pulls)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![BioC status](http://www.bioconductor.org/shields/build/release/bioc/HMMERutils.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/HMMERutils)
[![R-CMD-check-bioc](https://github.com/currocam/HMMERutils/actions/workflows/R-CMD-check-bioc.yaml/badge.svg)](https://github.com/currocam/HMMERutils/actions/workflows/R-CMD-check-bioc.yaml)
[![Codecov test coverage](https://codecov.io/gh/currocam/HMMERutils/branch/master/graph/badge.svg)](https://app.codecov.io/gh/currocam/HMMERutils?branch=master)
<!-- badges: end -->

The goal of `HMMERutils` is to provide a bunch of convenient functions to search for homologous sequences using the HMMER API, annotate them taxonomically, calculate physicochemical properties and facilitate exploratory analysis of homologous sequence data.

## Installation instructions

Not yet !

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then install `HMMERutils` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("HMMERutils")
```

And the development version from [GitHub](https://github.com/currocam/HMMERutils) with:

```{r 'install_dev', eval = FALSE}
BiocManager::install("currocam/HMMERutils")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example, eval = requireNamespace('HMMERutils'), warning=FALSE, message=FALSE}
library("HMMERutils")
fasta_2abl <- paste0(
    "MGPSENDPNLFVALYDFVASGDNTLSITKGEKLRVLGYNHNGEWCEAQTKNGQGW",
    "VPSNYITPVNSLEKHSWYHGPVSRNAAEYLLSSGINGSFLVRESESSPGQRSISL",
    "RYEGRVYHYRINTASDGKLYVSSESRFNTLAELVHHHSTVADGLITTLHYPAP"
)
data <- search_phmmer(seq = fasta_2abl, seqdb = "swissprot") %>%
    add_sequences_to_hmmer_tbl() %>%
    add_taxa_to_hmmer_tbl() %>%
    add_physicochemical_properties_to_HMMER_tbl()
```


```{r summary}
library(dplyr)
data %>%
    filter(hits.evalue > 10^-5) %>%
    distinct(hits.fullfasta, .keep_all = TRUE) %>%
    group_by(taxa.phylum) %>%
    summarise(
      n = n(),
      "Molecular.Weigth" = mean(properties.molecular.weight, na.rm = TRUE)        )
```



## Citation

Below is the citation output from using `citation('HMMERutils')` in R. Please
run this yourself to check for any updates on how to cite __HMMERutils__.

```{r 'citation', eval = requireNamespace('HMMERutils')}
print(citation("HMMERutils"), bibtex = TRUE)
```

Please note that the `HMMERutils` was only made possible thanks to many other R and bioinformatics software authors, which are cited either in the vignettes and/or the paper(s) describing this package.

## Code of Conduct

Please note that the `HMMERutils` project is released with a [Contributor Code of Conduct](http://bioconductor.org/about/code-of-conduct/). By contributing to this project, you agree to abide by its terms.

## Development tools

* Continuous code testing is possible thanks to [GitHub actions](https://www.tidyverse.org/blog/2020/04/usethis-1-6-0/)  through `r BiocStyle::CRANpkg('usethis')`, `r BiocStyle::CRANpkg('remotes')`, and `r BiocStyle::CRANpkg('rcmdcheck')` customized to use [Bioconductor's docker containers](https://www.bioconductor.org/help/docker/) and `r BiocStyle::Biocpkg('BiocCheck')`.
* Code coverage assessment is possible thanks to [codecov](https://codecov.io/gh) and `r BiocStyle::CRANpkg('covr')`.
* The [documentation website](http://currocam.github.io/HMMERutils) is automatically updated thanks to `r BiocStyle::CRANpkg('pkgdown')`.
* The code is styled automatically thanks to `r BiocStyle::CRANpkg('styler')`.
* The documentation is formatted thanks to `r BiocStyle::CRANpkg('devtools')` and `r BiocStyle::CRANpkg('roxygen2')`.

