---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# HMMERutils

<!-- badges: start -->
[![GitHub issues](https://img.shields.io/github/issues/currocam/HMMERutils)](https://github.com/currocam/HMMERutils/issues)
[![GitHub pulls](https://img.shields.io/github/issues-pr/currocam/HMMERutils)](https://github.com/currocam/HMMERutils/pulls)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![BioC status](http://www.bioconductor.org/shields/build/release/bioc/HMMERutils.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/HMMERutils)
[![R-CMD-check-bioc](https://github.com/currocam/HMMERutils/workflows/R-CMD-check-bioc/badge.svg)](https://github.com/currocam/HMMERutils/actions)
<!-- badges: end -->

The goal of `HMMERutils` is to provide convenient functions search for homologous sequences using the HMMER API, annotate them taxonomically, calculate physicochemical properties and facilitate exploratory analysis of homologous sequence data.

## Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then install `HMMERutils` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("HMMERutils")
```

And the development version from [GitHub](https://github.com/currocam/HMMERutils) with:

```{r 'install_dev', eval = FALSE}
BiocManager::install("currocam/HMMERutils")
```
## Example

This is a basic example showing how to post a request to HMMER, convert the information it returns into a tidy HMMERutils' DataFrame, as well as add taxonomic and physicochemical information inferred from the sequence. As an example, we will use the results of the PHMMER example search, where we searched for sequences homologous to ABL1 in the PDB. 

First, we post the request:
```{r, message=FALSE, warning=FALSE}
library("HMMERutils")
library(tidyverse)
library(Biobase)
ABL1_seq <- "MGPSENDPNLFVALYDFVASGDNTLSITKGEKLRVLGYNHNGEWCEAQTKNGQGWVPSNYITPVNSLEKHSWYHGPVSRNAAEYLLSSGINGSFLVRESESSPGQRSISLRYEGRVYHYRINTASDGKLYVSSESRFNTLAELVHHHSTVADGLITTLHYPAP"

## Downloaded files from HMMER
HMMER_ABL1_search <- search_phmmer(ABL1_seq,dbs = "pdb", verbose = FALSE)
HMMER_ABL1_search
```

The different files that HMMER makes available to us, as well as the information about the sequences are now contained in `HMMER_ABL1_search`. Let's see the metadata of the first 6 variables of this `AnnotatedDataFrame`:

```{r, message=FALSE, warning=FALSE}
head(varMetadata(HMMER_ABL1_search))
```

We can then extract this information into a tidy DataFrame, preprocess the data and add more information to it. The basic workflow is as follows:

```{r, message=FALSE, warning=FALSE}
HMMER_ABL1_data <- HMMER_ABL1_search %>%
  extract_from_HMMER_data_tbl() %>% # extract the information into a DataFrame
  add_fullseq_to_HMMER_tbl(HMMER_ABL1_search$fullfasta.url)%>% #add sequences
  add_physicochemical_properties_to_HMMER_tbl() #calculate theoretical physical and chemical properties
HMMER_ABL1_data
```

 Let's see the metadata of the first 15 variables of this `AnnotatedDataFrame`:
 
```{r, message=FALSE, warning=FALSE}
head(varMetadata(HMMER_ABL1_data),15)
```

## Citation

Below is the citation output from using `citation('HMMERutils')` in R. Please
run this yourself to check for any updates on how to cite __HMMERutils__.

```{r
print(citation("HMMERutils"), bibtex = TRUE)
```

Please note that the `HMMERutils` was only made possible thanks to many other R and bioinformatics software authors, which are cited either in the vignettes and/or the paper(s) describing this package.

## Code of Conduct

Please note that the `HMMERutils` project is released with a [Contributor Code of Conduct](http://bioconductor.org/about/code-of-conduct/). By contributing to this project, you agree to abide by its terms.

## Development tools

* Continuous code testing is possible thanks to [GitHub actions](https://www.tidyverse.org/blog/2020/04/usethis-1-6-0/)  through `r BiocStyle::CRANpkg('usethis')`, `r BiocStyle::CRANpkg('remotes')`, and `r BiocStyle::CRANpkg('rcmdcheck')` customized to use [Bioconductor's docker containers](https://www.bioconductor.org/help/docker/) and `r BiocStyle::Biocpkg('BiocCheck')`.
* Code coverage assessment is possible thanks to [codecov](https://codecov.io/gh) and `r BiocStyle::CRANpkg('covr')`.
* The [documentation website](http://currocam.github.io/HMMERutils) is automatically updated thanks to `r BiocStyle::CRANpkg('pkgdown')`.
* The code is styled automatically thanks to `r BiocStyle::CRANpkg('styler')`.
* The documentation is formatted thanks to `r BiocStyle::CRANpkg('devtools')` and `r BiocStyle::CRANpkg('roxygen2')`.

For more details, check the `dev` directory.

This package was developed using `r BiocStyle::Biocpkg('biocthis')`.


