---
title: "Introduction to HMMERutils: using PHMMER with the Sars-CoV-2 Reference Proteome"
author: 
  - name: Francisco de Borja Campuzano Jiménez
    affiliation:
    - Universidad de Granada
    email: campuzanocurro@gmail.com
  - name: María Coral del Val Muñoz
    affiliation:
    - Universidad de Granada
    email: delval@ugr.es   
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('HMMERutils')`"
vignette: >
    %\VignetteIndexEntry{Introduction to HMMERutils}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL 
)
```


```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Track time spent on making the vignette
startTime <- Sys.time()

## Bib setup
library("RefManageR")

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    Biostrings = citation("Biostrings")[1],
    Peptides = citation("Peptides")[1],
    Biobase = citation("Biobase")[1],
    dplyr = citation("dplyr")[1],
    ggplot2 = citation("ggplot2")[1],
    graphics = citation("graphics")[1],
    magrittr = citation("magrittr")[1],
    methods = citation("methods")[1],
    purrr = citation("purrr")[1],
    RCurl = citation("RCurl")[1],
    rlang = citation("rlang")[1],
    stringr = citation("stringr")[1],
    tibble = citation("tibble")[1],
    tidyr = citation("tidyr")[1],
    XML = citation("XML")[1],
    HMMERutils = citation("HMMERutils")[1]
)
```

# Basics

## Install `HMMERutils`

`R` is an open-source statistical environment which can be easily modified to enhance its functionality via packages. `r Biocpkg("HMMERutils")` is a `R` package available via the [Bioconductor](http://bioconductor.org) repository for packages. `R` can be installed on any operating system from [CRAN](https://cran.r-project.org/) after which you can install `r Biocpkg("HMMERutils")` by using the following commands in your `R` session:

```{r "install", eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("HMMERutils")

## Check that you have a valid Bioconductor installation
BiocManager::valid()
```

## Required knowledge

`r Biocpkg("HMMERutils")` is based on many other packages and in particular those that have implemented the infrastructure necessary to deal with protein sequence data and generate ` AnnotatedDataFrames`. Therefore, note that the `r Biocpkg("Biostrings")` libraries are used to read, write and handle protein sequences and `r Biocpkg("Biobase")` is used to generate the annotated DataFrames that the user works with.  The content of the ` AnnotatedDataFrames` have been designed for this purpose. 


## Asking for help

As package developers, we try to explain clearly how to use our packages and in which order to use the functions. But `R` and `Bioconductor` have a steep learning curve so it is critical to learn where to ask for help. The blog post quoted above mentions some but we would like to highlight the [Bioconductor support site](https://support.bioconductor.org/) as the main resource for getting help: remember to use the `HMMERutils` tag and check [the older posts](https://support.bioconductor.org/t/HMMERutils/). Other alternatives are available such as creating GitHub issues and tweeting. However, please note that if you want to receive help you should adhere to the [posting guidelines](http://www.bioconductor.org/help/support/posting-guide/). It is particularly critical that you provide a small reproducible example and your session information so package developers can track down the source of the error.

## Citing `HMMERutils`

We hope that `r Biocpkg("HMMERutils")` will be useful for your research. Please use the following information to cite the package and the overall approach. Thank you!

```{r "citation"}
## Citation info
citation("HMMERutils")
```

# Quick start to using to `HMMERutils`

```{r, prerequisites, message=FALSE, warning=FALSE}
library(HMMERutils)
library(Biostrings)
library(Biobase)
library(tidyverse)
library(ggplot2)
```

The objective of this article is to show the main functionalities of the utilsHMMER library using a case study. We will search for homologous sequences for the whole SARS-CoV-2 reference proteome.

## Download the SARS-CoV-2 reference proteome 

First, let’s read the reference proteome for the Sars_CoV-2. To avoid problems with future versions, we have already downloaded this proteome from Uniprot and it is included as sample data for this library. Then we will read the compressed file directly from UniProt with the `readAAStringSet` function from the `r Biocpkg("Biostrings")` library.

```{r, warning=FALSE}
sars.cov.fasta <- system.file("extdata/UP000000354_694009.fasta", package = "HMMERutils")%>%
  readAAStringSet()
```

## Preprocess sequences and select case studies

Now, we have an `AAStringSet` stored in the variable `sars.cov.fasta` and we can access the sequences it contains using the `$` operator and the sequence header. However, headers usually contain, as in this case, a lot of information that is not useful right now such as descriptions. `r Biocpkg("HMMERutils")` has a very convenient function for this purpose which is parse. Note that we use the `name` function to access the sequence names and the `%>%` operator to pass the result of one function/argument to the other in sequence

```{r, message=FALSE, warning=FALSE}
new.headers <- sars.cov.fasta %>%
  names() %>%
  parse_FASTA_sequence_identifiers()
new.headers
```

Now, we can assign these new headers to the `AAStringSet`:

```{r, message=FALSE, warning=FALSE}
names(sars.cov.fasta) <- new.headers
```

We can take a look at the contents of the fasta file by "showing" it:

```{r, message=FALSE, warning=FALSE}
sars.cov.fasta
```

Now, we are going to select a series of proteins that are of interest to us for this case study:

```{r, message=FALSE, warning=FALSE}
VEMP_SARS <- sars.cov.fasta$VEMP_SARS #Envelope small membrane protein
NS8A_SARS <- sars.cov.fasta$NS8A_SARS #ORF8a protein
NS8B_SARS <- sars.cov.fasta$NS8B_SARS #ORF8b protein
NS7B_SARS <- sars.cov.fasta$NS7B_SARS #Protein non-structural 7b
Y14_SARS <- sars.cov.fasta$Y14_SARS #Uncharacterized protein
```

All these proteins have in common to be small in size (it will speed up some computational times) and represent different possibilities: we have from membrane proteins (which we would expect to show homology with other membrane proteins), proteins similar to each other (NS8A_SARS and NS8B_SARS), non-structural proteins and uncharacterized proteins. 

## Proteins homologous to VEMP_SARS using PHMMER

Before exemplifying how to use HMMERutils to perform several searches simultaneously (usually against several databases) or to use other algorithms, let's perform a basic search against a single database using `phmmer`, the default algorithm. 

### Post request to HMMER server 

Now, we can start posting requests to HMMER through its API, using the `search` family of functions. Let's start with an example: let's search for proteins homologous to VEMP_SARS (a envelope small membrane protein) against the Swissprot database. Notice that we have indicated as sequence the `AAString` we defined before (although it could also be a character) and we have given the parameter `verbose = FALSE`, indicating that we do not want the progress of the search to be printed because it would take too much space in this document.

```{r, message=FALSE, warning=FALSE}
HMMER_data_tbl <- search_phmmer(
  seqs = VEMP_SARS,
  dbs = "swissprot",
  verbose = FALSE)
```

### AnnotatedDataFrame returned by "search" function family

Let's inspect the content of the new variable we have created:

```{r, message=FALSE, warning=FALSE}
HMMER_data_tbl
```
As we can see, this is an `AnnotatedDataFrame`, an class object of the `r Biocpkg("Biobase")` library that allows us to store a DataFrame together with its metadata. This is the way in which the `search_` family of functions will return the results. We can also see that it has 1 row (since it contains the results of a single search) and 17 variables. Let's see what these variables and their descriptions are by accessing the metadata using the `VarMetadata` function or the `@` operator.

```r
varMetadata(HMMER_data_tbl)
```

```{r, echo=FALSE}
varMetadata(HMMER_data_tbl) %>%
  kableExtra::kbl(caption = "Metadata associated with each of the searches performed in HMMER through HMMERutils.")
```
As can be seen in the table above, we have different types of variables that may be of interest in `VEMP_SARS.homologous.PDB`. 

#### Files available for download and urls

On the one hand, we have a series of parameters that refer to our own search (and that in this case are not useful, since we have performed a single search) and that contain a series of urls. Variables starting with "url" contain the download links of all download files that will be available for a limited time on the HMMER web server. Also, in `score.url` is the web page with the HMMER results (the same one you would get to if you were using the web interface), which you can access to consult the results from there. 

The purpose of this library is not to be an intermediary with HMMER web server. However, in case you need to post a large number of searches, it could be used to automate sending the search parameters and return a table with all the links that might be of interest to you!

To display such a table we have several options. The simplest option is to use the `tidyverse` metalibrary. However, in order to do this we must first separate the `AnnotatedDataFrame` data from its metadata. To do this, we could use the `pData()` function or the `@` operator. Then, we select only the parameters ending with `url` using the syntax shown below (which will be useful later as well) and that we transpose the table to make it easier to read in this document.

``` r
 HMMER_data_tbl@data%>%
   select(ends_with("url")) %>%
   t()
#pData(VEMP_SARS.homologous.PDB) %>%
#  select(ends_with("url")) %>%
#  t()
```

```{r, echo=FALSE}
pData(HMMER_data_tbl) %>%
  select(ends_with("url")) %>%
  t()%>%
  kableExtra::kbl(caption = "Links to files with information about the search performed in HMMER. Note that they have probably already expired. ")
```

As you may have guessed, we can use the `readAAStringSet` function to, just as we read the Fasta file containing the Sars-CoV-2 Reference Proteome, via a url, we can use it to download and read the HMMER files. 

```{r, message=FALSE, warning=FALSE, eval=FALSE}
readAAStringSet(HMMER_data_tbl$fullfasta.url)
```

Likewise, we could read the alignments of these sequences to the HMM (Hidden Markov Model) generated by HMMER using the function `readAAMultipleAlignment` and indicating the format:

```{r, message=FALSE, warning=FALSE, eval=FALSE}
readAAMultipleAlignment(HMMER_data_tbl$alignedfasta.url)
```

#### Information about the sequences found

On the other hand, we have the `stats`, `hits` and `domains` parameters, which contain the hashes returned by HMMER for each search. These are actually nested DataFrames which in turn contain another set of parameters. Let's see the content of the nested DataFrame `stats`, although, as we will discuss later, there is a more convenient way to deal with this information. 

Note that, being a nested DataFrame, it is not enough to make use of the `$` operator, but we also have to use `[[1]]` to indicate that we want the first element of the list or, otherwise, we would get a list with, in this case, a single element that would be the DataFrame instead of the DataFrame itself. Although this may seem awkward now, it is necessary to maintain consistency when several searches are made instead of one (as we will see in the next section). 

``` r
t(HMMER_data_tbl$stats[[1]])
```

```{r, echo=FALSE}
t(HMMER_data_tbl$stats[[1]]) %>%
  kableExtra::kbl(caption = "Content of the hash stats, with internal HMMER information and variables about the search itself and not about its content. ")
```
Actually, most of the parameters contained in `stats` are for internal use and we are only interested in a small part of them.  For example, by studying the table above, we can know that HMMER found 40 hits (`nhits`) out of a total of 565928 sequences available (`Z`) in the PDB and that a single HMM (Hidden Markov Model) (`nmodels`) has been used in the search (built from our input sequence since we are using PHMMER algorithm). In case you are interested in learning more about how the results data are structured, we invite you to read the appendix of the [HMMER web server](https://hmmer-web-docs.readthedocs.io/en/latest/appendices.html). 

We are not going to repeat the process for `hits` and `domains` because there is a much better way to query the information returned by HMMER and that is the `extract_from_HMMER_data_tbl` function. We show this possibility to point out that, in case the functionalities of use that we are going to show next and that are the core of this library are not useful for your application, you can access all the information as returned by HMMER in this way instead of parsing the JSON or XML file directly. 

### Extracting a Tidy Dataframe

Let's extract the contents of `hits` and `domains` to a tidy DataFrame using the `extract_from_HMMER_data_tbl` function. From now on, it will be this object that we will work with and not the one containing the search results. 

```{r, message=FALSE, warning=FALSE}
HMMER_tidy_tbl <- extract_from_HMMER_data_tbl(HMMER_data_tbl)
```

As you can see, the result is, again, an `AnnotatedDataFrame` that in this case has as many rows as domains have been identified (keep this in mind in case, as is likely, you want to work by identified sequences and not necessarily the identified domains within these sequences). In this case, there are a total of 40 rows, i.e. the number of domains coincides with the number of sequences, so that only one domain per sequence has been identified although, especially in larger proteins, this is not usually the case.  

```{r, message=FALSE, warning=FALSE}
HMMER_tidy_tbl
```

Let's inspect the parameters it contains and their meaning by inspecting the table with the metadata, as we did before:

```{r, message=FALSE, warning=FALSE}
#metadata.VEMP_SARS <- HMMER_tidy_tbl@varMetadata
metadata.VEMP_SARS <- varMetadata(HMMER_tidy_tbl)
```

```r
metadata.VEMP_SARS
```

```{r, echo=FALSE}
metadata.VEMP_SARS %>%
  kableExtra::kbl(caption = "The metadata of the full sequence and domains found by HMMER. ")
```

### Data preprocessing

#### E-values

First, let us begin by studying the` e-values` obtained. Then, first of all, we should look at the E-value of the full sequence to see if it is lower than our significance threshold. This can be done as follows. Note that we use the `@data` operator, because we want to access and overwrite the data, not the metadata. 

```{r, message=FALSE, warning=FALSE}
threshold <-  0.01
pData(HMMER_tidy_tbl) <- pData(HMMER_tidy_tbl) %>%
  filter(hits.evalue <threshold)
```


As you can see in previous table, there are several types of ` e-values`, some referring to the sequence and others referring to the domains. But what do we mean by domains? According to the HMMER manual:

> A target sequence that reaches this point is very likely to contain one or more significant matches to the profile. These matches are referred to as “domains”, since the main use of HMMER has historically been to match profile HMMs from protein domain databases like
Pfam, and one of HMMER’s strengths is to be able to cleanly parse a multidomain target sequence into its multiple nonoverlapping hits to the same domain model.

It is of special interest to study carefully the domain `E-value` when it happens that, although the full sequence `E-value` is good, none of the domain `E-value` reach the significance threshold. In that case, what is happening is that those weak hits, none of which is good enough on its own, add up to raise the sequence to a high score. This may happen because the sequence contains several weak homologous domains, or it may contain a repetitive sequence that is hitting by chance.

HMMERutils has a function to display all the `E-values` of the sequences. This function `hmmer_evalues_cleveland_dot_plot`. In this plot the $-log(E-value)$ is represented on the $x$ axis and on the $y$ axis each of the identified sequences is represented one at a time. The green dots represent the E-values of the fullsequence and the red dots the E-values of each of the domains. 

```{r, message=FALSE, warning=FALSE, fig.cap= "Cleveland dot plot of the E-values of each of the domains (colored in red) and of the complete sequences (colored in green) homologous to VEMP_SARS. "}
hmmer_evalues_cleveland_dot_plot(HMMER_tidy_tbl, threshold = threshold)
```
By looking at it, you can see at a glance, first, how many domains have been identified for each sequence, and then how these contribute to the E-value of the fullsequence and how significant they are relative to the fullsequence. In this case, as can be seen, only one domain per sequence has been found (probably because they are very small) but we will see a case later where more than one domain are found. 

#### Removing redundant sequences
Now, let's filter the sequences to keep only non-redundant sequences. First, we have to add the sequences to the `AnnotatedDataFrame` by downloading the fullfasta file.  To do this, first download the compressed file as a binary, then unzip it, read and add it as a new column using the `add_fullseq_to_HMMER_tbl` function.


To do this we will use the `add_fullseq_to_HMMER_tbl` function, giving as arguments the AnnotatedDataFrame and a character vector with the addresses to the file /fullfastafastafiles. Note that this address can point to a file downloaded to your computer (if you have downloaded the files from the page) or the URL.

```{r, message=FALSE, warning=FALSE}
HMMER_tidy_tbl <- HMMER_tidy_tbl %>%
  add_fullseq_to_HMMER_tbl()
```
Actually, this process is not necessary if you are using a Unix-like operating system, since, unlike Windows, it does not distinguish between text and binary files. In case you are using a Unix-like operating system, you can simply run the following chunk:

```{r, eval=FALSE}
HMMER_tidy_tbl <- HMMER_tidy_tbl %>%
  add_fullseq_to_HMMER_tbl()
```


Now, we can remove non redundant sequences using the `distinc` function. Note that we specify the parameter `.keep_all = TRUE` because we want to keep all columns. 

```{r, message=FALSE, warning=FALSE}
pData(HMMER_tidy_tbl) <- pData(HMMER_tidy_tbl) %>%
  distinct(hits.fullfasta, .keep_all = TRUE)
```
Let's see how the `AnnotatedDataFrame` has now changed.

```{r, message=FALSE, warning=FALSE}
HMMER_tidy_tbl
```
### Data exploration

#### Taxonomic distribution

Now that we have selected our 23 sequences of interest, we can explore the nature of these sequences. One aspect of great interest is the taxonomic distribution of these sequences. We can consult it because we have the `hits.taxid` parameter, which identifies taxonomic species. We proceed to taxonomically annotate the results using the `add_taxa_to_HMMER_tbl` function

```{r, message=FALSE, warning=FALSE}
HMMER_tidy_tbl <- HMMER_tidy_tbl %>%
  add_taxa_to_HMMER_tbl(mode = "local")
```
Note that we have indicated the parameter mode = "local" to note that we don't want to use remote resources. Using remote resources can be slow, we advise you, if the amount of sequences is considerable, to use the "local" option to install (only once) and then use a local database. Let us see what has happened in our `AnnotatedDataframe`.
```{r, message=FALSE, warning=FALSE}
HMMER_tidy_tbl
```

As you can see, we have added 15 new variables (the number will depend on the taxonomic categories available). Let's see which variables they are. To display only the ones we have just added we will use the following syntax (last 15 rows):

``` r
HMMER_tidy_tbl %>%
  varMetadata() %>%
  slice_tail(n=15)
```

```{r, echo=FALSE}
HMMER_tidy_tbl %>%
  varMetadata() %>%
  slice_tail(n=15) %>%
  kableExtra::kbl(caption = "Taxonomic variables metadata")
```

Now, we can use all the metallibrary of `tidyverse` and `ggplot2` to explore the data and make graphs. Let's see an example of what could be done:

##### Example: how to plot taxonomic information using ggplot2

```{r, message=FALSE, warning=FALSE, fig.cap="Histogram of VEMP_SARS homologous sequences E-values according to taxonomic subgenus"}
HMMER_tidy_tbl %>%
  pData() %>%
  ggplot(aes(x = hits.evalue, fill = taxa.subgenus)) + 
  geom_histogram(binwidth = 0.0001)
```
#### Calculating the identity of pairwise alignments

Based on the  previous figure, it would be reasonable to think that there would be a greater diversity in the Embecovirus sequences, since there is a greater variety in the E-values. To examine this in more detail we can align all the sequences in pairs and then calculate the percentages of identity for each one Using the `pairwise_alignment_sequence_identity` function.

When calling this function, we give it as parameters the variable `seqs` containing the complete sequences (we could give it the sequences of the domains), the type of alignment and the type of sequence identity we are going to use. In this case we will use the default options: `aln_type = "global"` (align whole strings with end gap penalties by using the Needleman-Wunsch algorithm) and `pid_type = "PID1"` ( $PID1 = 100 \cdot \frac{\text{identical positions}}{\text{aligned positions + internal gap positions}}$). In addition, we could use the `allow_parallelization` parameter to speed up the calculation. The option "multisession" option will generate several Rstudio sessions on your computer, while the "multicore" option, is considered unsafe to use in an Rstudio session (although you could use it in a script). 

```{r, message=FALSE, warning=FALSE}
pid <- pairwise_alignment_sequence_identity(seqs = HMMER_tidy_tbl$hits.fullfasta,
                                     aln_type = "global",
                                     pid_type = "PID1",
                                     #allow_parallelization = "multisession"
                                     )
```

Now, we can use 2 HMMERutils functions to view the results. We can see the distribution of these identities:  

```{r, message=FALSE, warning=FALSE, fig.cap= "Histogram of sequence identity by pairs of sequences homologous to VEMP_SARS."}
pairwise_sequence_identity_histogram(pid)
```
And, in order to compare this information with other parameters, a heatmap. Note that we use the `taxa.subgenus` column to annotate the heatmap, for which we use the `annotation` argument and the `$` operator. 

```{r, message=FALSE, warning=FALSE, fig.cap= "Heatmap of sequence identity by pairs of sequences homologous to VEMP_SARS."}
pairwise_sequence_identity_heatmap(pid,annotation = HMMER_tidy_tbl$taxa.subgenus)
```
As we can now see, the *Embecovirus* sequences form 2 groups with a high sequence identity with each other and with themselves. In addition, there are 2 sequences that present a very low percentage of sequence identity with the rest of the sequences of their subgenus, but high with the *Tegacovirus* sequences. 

#### Physicochemical properties

We can study what the sequences we have obtained are like in more detail by calculating physicochemical properties based on their primary sequence. To do this, we can use the `add_physicochemical_properties_to_HMMER_tbl` function:
```{r, message=FALSE, warning=FALSE}
HMMER_tidy_tbl <- HMMER_tidy_tbl %>% 
  add_physicochemical_properties_to_HMMER_tbl()
```

Let us see what has happened in our `AnnotatedDataframe`.
```{r, message=FALSE, warning=FALSE}
HMMER_tidy_tbl
```

As you can see, we have added 18 new variables. In this case it is almost essential to examine the metadata of the new parameters, in order to know what we are talking about. We will use the following syntax (last 18 rows):

``` r
HMMER_tidy_tbl %>%
  varMetadata() %>%
  slice_tail(n = 18)
```

```{r, echo=FALSE}
HMMER_tidy_tbl %>%
  varMetadata() %>%
  slice_tail(n=18) %>%
  kableExtra::kbl(caption = "Metadata of the variables on the theoretical physicochemical properties calculated with the primary structure. ")
```

##### Example: how to perform a PCA on the physicochemical properties of sequences using factoextra   

Let's see an example of how to use these theoretical parameters that we have just calculated. To perform a PCA we must first select only the columns that contain these properties (and we do it taking advantage of the fact that they all begin with "properties"), scale them (`scale = TRUE`) and perform the PCA with the `prcomp` function.

```{r, message=FALSE, warning=FALSE}
pca.properties <- pData(HMMER_tidy_tbl) %>%
  select(starts_with("properties")) %>%
  prcomp(scale = TRUE)
```

Now, using the `factoextra` library, we are going to project the homologous sequences to VEMP_SARS on the 2 main components to color them according to the taxonomic subgenus.  

```{r, message=FALSE, warning=FALSE, fig.cap="Principal component analysis on proteins homologous to VEMP_SARS. Sequence projection."}
library(factoextra)
fviz_pca_ind(pca.properties, geom.ind = "point", 
             col.ind = HMMER_tidy_tbl$taxa.subgenus, 
             axes = 1:2, 
             pointsize = 1.5) 
```
Now, we can see how, although we would have thought otherwise if we had paid attention exclusively to the percentage of sequence identity, Embecovirus sequences are not so different physico-chemically, even though their sequence is. We may also be interested in knowing which variables are most involved. To do this, we are going to make a biplot:
```{r, message=FALSE, warning=FALSE, fig.cap= "Principal component analysis on proteins homologous to VEMP_SARS. Biplot of variables and sequences."}
fviz_pca(pca.properties, geom.ind = "point", 
             col.ind = HMMER_tidy_tbl$taxa.subgenus, 
             axes = 1:2, 
             pointsize = 2) 
```
## Proteins homologous to Y14_SARS in SwissProt and Reference Proteomes using PHMMER

So far, we have been working with the results of a single search. This case study has allowed us to describe in depth most of the utilities of this library. However, the most common scenario is that we want to work with the results of several searches. Let's try performing a search with PHMMER for the sequence of Y14_SARS, an uncharacterized protein, against SwissProt and Ensembl. 

For this example, moreover, we will not go step by step, but will show how HMMERutils can perform all of the above operations in a few lines of code. In case you find the keyword "uniprotrefprot" unintuitive, which it certainly is, you can refer to the information regarding the `search_phmmer` function using ?search_phmmer for a complete listing of options. 
```{r}
HMMER_data_Y14_SARS <- Y14_SARS %>%
  search_phmmer(
  dbs = c("swissprot", "uniprotrefprot"),
  verbose = FALSE)
HMMER_data_Y14_SARS
```
Note that, now, the number of searches performed is 2. Let's download both files:


Now, using the pipe operator, we are going to extract the information in a table, read the sequences, annotate them taxonomically and calculate the physicochemical properties of the sequences. 

```{r}
HMMER_Y14_SARS <- HMMER_data_Y14_SARS %>%
  extract_from_HMMER_data_tbl() %>%
  add_fullseq_to_HMMER_tbl() %>%
  add_taxa_to_HMMER_tbl() %>%
  add_physicochemical_properties_to_HMMER_tbl()
HMMER_Y14_SARS
```
Again, remember that, in case you are using a Unix-like operating system, it is not necessary to download the files first, and the following simplified procedure would be equally valid: 

```{r, eval=FALSE}
HMMER_Y14_SARS <- HMMER_data_Y14_SARS %>%
  extract_from_HMMER_data_tbl() %>%
  add_fullseq_to_HMMER_tbl() %>%
  add_taxa_to_HMMER_tbl() %>%
  add_physicochemical_properties_to_HMMER_tbl()
HMMER_Y14_SARS
```

## Proteins homologous to various sequences and target databases using PHMMER

Finally, let's perform an example by searching for homologous sequences of several sequences against several databases. We will take advantage of this case study to exemplify how the data could be treated when a large number of sequences are found. 

We will work with the sequences NS8A_SARS, NS8B_SARS and NS7B_SARS. First, we must save the sequences of interest in a list. Notice that we save the sequences in a list instead of in a vector because, being AAStringSet, the default behavior would be to concatenate the sequences instead of saving them separately. 
```{r}
NS8A_SARS <- sars.cov.fasta$NS8A_SARS #ORF8a protein
NS8B_SARS <- sars.cov.fasta$NS8B_SARS #ORF8b protein
NS7B_SARS <- sars.cov.fasta$NS7B_SARS #Protein non-structural 7b
NSXX_SARS <- list(NS8A_SARS, NS8B_SARS, NS7B_SARS)
```

As we can see, by indicating 3 searches against 2 databases we have performed the corresponding 6 searches. 

```{r}
HMMER_data_NSXX_SARS <- NSXX_SARS %>%
  search_phmmer(
  dbs = c("swissprot", "uniprotrefprot"),
  verbose = FALSE)
HMMER_data_NSXX_SARS
```

And now let's extract the data and add the fullfasta sequences. 

```{r}
HMMER_NSXX_SARS <- HMMER_data_NSXX_SARS %>%
  extract_from_HMMER_data_tbl() %>%
  add_fullseq_to_HMMER_tbl() 
HMMER_NSXX_SARS
```

##### Example: how to manipulate dataframe with Tidyverse

We could fall into the error of thinking that we have obtained 204 sequences from these 6 searches. However, there are a number of factors to take into account: 

- We are searching in redundant databases, so it is to be expected that there will be, therefore, redundant sequences.  
- There may be redundant sequences of different species. 
- We have not yet filtered out non-significant sequences according to their E-Value.
- We can have sequences that correspond to fragments.
- And, as mentioned above, the number of rows in the table we have just created corresponds to each of the homologous domains identified for each sequence, which may have several. 


All these manipulation tasks will be specific to each case but, probably, the easiest way to carry them out is using `Tidyverse.`  Let's see how many full sequences we have obtained by building a pipeline step by step:

First, we want to exclude sequences with an E-Value greater than 0.01: 
``` r
HMMER_NSXX_SARS %>%
  pData() %>%
  filter(hits.evalue <0.01)
```
To remove entries that refer more than once to the same sequence, we have several options. For example, in the following way, we will keep the sequences that are different in their primary and secondary accession (in case they share the primary). Note that we indicate `.keep_all = TRUE` in order not to delete columns. 
``` r
HMMER_NSXX_SARS %>%
  pData() %>%
  distinct(hits.acc, hits.acc2, .keep_all = TRUE)
```

Also note that, having filtered by acc and acc2, we are actually removing every domain row except the first one In case we want to keep the domains, i.e. still have one row per domain instead of one row per sequence, we can add, for example the `domains.ienv` parameter (envelope start position).
``` r
HMMER_NSXX_SARS %>%
  pData() %>%
  distinct(hits.acc, hits.acc2, domains.ienv, .keep_all = TRUE)
```

As we have shown, Tidyverse is very flexible and allows for really complex combinations. However, in most cases, we will only be interested in filtering by the full sequence (remove 100% of redundant sequences). Note that, to alter the `AnnotatedDataframe` data, we must use the `pData` function or the `@` operator twice: on the left side of the assignment, to indicate what we want to change, and on the right side, to access this information.  

```{r}
pData(HMMER_NSXX_SARS) <- HMMER_NSXX_SARS%>%
  pData() %>%
  filter(hits.evalue <0.01) %>%
  distinct(hits.fullfasta, .keep_all = TRUE)
```

As you can see, of the 204 domains that showed homology, we obtained 43 entries, corresponding to 43 non-redundant sequences homologous to the sequences of interest.

## Argonaute proteins homologous using hmmsearch

So far, we have only used PHMMER, since it is the most used and default algorithm. However, HMMERutils allows to work with all HMMER algorithms. Let's see how it can be used to perform a search using HMMSEARCH. This algorithm performs a sequence search in the same databases as PHMMER with the difference that, instead of building a hidden Markov model with a single sequence, it does it with an alignment. 

For this example we will use an alignment of the PIWI domain of PFAM, which belongs to the Argonaute proteins (catalytic components of the RNA-induced silencing complex (RISC)) obtained from PFAM. With this alignment, we will perform a search with HMMSEARCH against the PDB, imagining, for example, that our ultimate intention is to find protein structures to work with. 


First, let's read the alignment:
```{r}
piwi_seed_aln_path <- system.file("extdata/piwi.afa",package = "HMMERutils")
piwi_alignment <- readAAMultipleAlignment(piwi_seed_aln_path)
piwi_alignment
```

And now, let's do the search. 

```{r}
HMMER_piwi <- search_hmmsearch(
  alns = piwi_alignment,
  dbs = "pdb",
  verbose = FALSE) %>%
  extract_from_HMMER_data_tbl()
HMMER_piwi
```
Let's look at the PDBs you have found:
```{r}
HMMER_piwi$hits.name
```

## PIWI domain identification  using hmmscan

Another HMMER algorithm that can be used through HMMERutils is HMMSCAN. This algorithm performs a search of a protein sequence against an HMM database and can be used to identify domains in our sequence. 

To test it we are going to make use of a fragment of the PIWI domain which we have obtained by joining pieces of different Argonaute proteins:


```{r}
frankestein_piwi <- paste0(
  "YGSIKRVCNTMLGVPSQCAISKHILQS","SDQYLTNVLLKINSKLGGIN",
  "QYLTNLLLKINAKLGGLN", "SRIVADKITNKYLVDQPTM")
```
Now, let's perform the search against 2 HMM databases, `PFAM` and `tigrfam`

```{r}
HMMER_frankestein_piwi <- search_hmmscan(
  seqs = frankestein_piwi,
  dbs = c("pfam", "tigrfam"),
  verbose = FALSE) %>%
  extract_from_HMMER_data_tbl()
HMMER_frankestein_piwi
```
As expected, the search returned only 2 results and both correspond to the Piwi domain (2 results because we searched 2 redundant databases).
```{r}
HMMER_frankestein_piwi$hits.name
```
## Reading HMMER XML files

Finally, there is the possibility to read XML files downloaded directly from the HMMER website.  Let's see an example: 

```{r}
xml_example_path <- system.file("extdata/xml_example.xml", package = "HMMERutils")
HMMER_read_from_file <- read_hmmer_from_xml(xml_example_path) %>%
  extract_from_HMMER_data_tbl()
HMMER_read_from_file
```


# Reproducibility

Code for creating the vignette

```{r createVignette, eval=FALSE}
## Create the vignette
library("rmarkdown")
system.time(render("HMMERutils.Rmd", "BiocStyle::html_document"))

## Extract the R code
library("knitr")
knit("HMMERutils.Rmd", tangle = TRUE)
```

Date the vignette was generated.

```{r reproduce1, echo=FALSE}
## Date the vignette was generated
Sys.time()
```

Wallclock time spent generating the vignette.

```{r reproduce2, echo=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```



# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
with `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])` and `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`.

```{r vignetteBiblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
PrintBibliography(bib, .opts = list(hyperlink = "to.doc", style = "html"))
```
